Towards atomistic resolution structure of phosphatidylcholine headgroup and glycerol backbone at different ambient conditions\footnote{Publication about results presented in the NMRlipids project}

### 2015-05-14 HGmodelMANUSCRIPT/HGmodel_ACStemplate.tex
Careful reading of the cholesterol part.

\subsubsection{Cholesterol-containing phospholipid bilayer}
Phospholipid--cholesterol interactions have been widely studied with theoretical~\cite{huang99,zhu07,rog09,alwarawrah12} and
experimental methods~\cite{brown78,marsh10,ferreira13,marsh13}, since cholesterol is abundant in biological membranes.
It has been suggested to be an important player, for example, in domain formation~\cite{simons04,somerharju09}.
It is widely agreed that cholesterol orders lipid acyl tails thus decreasing the area per molecule (condensing effect),
however, the influence of cholesterol on the lipid headgroup and glycerol backbone are still debated~\cite{huang99,simons04,somerharju09}.
For example, it has been suggested that the surrounding lipids shield cholesterol from interactions with water by 
reorienting their headgroups (``umbrella model'')~\cite{huang99} or that cholesterol acts as a spacer for the headgroups thus increasing 
their entropy and dynamics (``superlattice model'')~\cite{somerharju09}. 
%\todo{It has been suggested that we should use direct quotations from original papers here for clarity.}
Both of these suggestions have been supported
by molecular dynamics simulations~\cite{zhu07,alwarawrah12}, and other simulations suggest specific
interactions between the glycerol backbone and cholesterol~\cite{rog09}. However,
in the above mentioned studies, the glycerol backbone and choline headgroup behaviour
as a function of cholesterol content has not been compared to experiments.

The choline headgroup and glycerol backbone order parameters for POPC measured by $^1$H-$^{13}$C NMR~\cite{ferreira13} and DPPC choline order parameters 
measured by $^{2}$H NMR~\cite{brown78} are shown in Fig.~\ref{ordPchol} as a function of cholesterol content.
The agreement between different experimental results is again very good, showing only very modest changes in 
the choline order parameters as a function of cholesterol content. It should be noted, however, that very small
changes are measurable with high resolution $^{2}$H NMR experiments
and cholesterol causes a measurable increase in the $\beta$ order parameter and a forking in the $\alpha$ order
parameter~\cite{brown78}. These effects are, however, so small that they are barely visible on the scale used in Fig.~\ref{ordPchol}.
Further, the effects of cholesterol on the glycerol backbone order parameters for POPC from $^1$H-$^{13}$C NMR experiment~\cite{ferreira13} 
are in good agreement with the results for the phosphatidylethanolamine (PE) measured by $^{2}$H NMR~\cite{ghosh82}.
These results further support the idea that the glycerol backbone structural behaviour is independent of the
headgroup composition~\cite{gally81} and that the headgroup structure is largely independent of the acyl chain region content unless
charges are present~\cite{scherer87}.
\begin{figure}[]
  \centering
  \includegraphics[width=7.3cm]{../DATAreportediINblog/cholesterolization_lp.pdf}
    \includegraphics[width=7.3cm]{../DATAreportediINblog/cholesterolization_smooth.pdf}
  \caption{\label{ordPchol}
    The effect of cholesterol content on the glycerol backbone and choline order parameters in experiments~\cite{brown78,ferreira13} and simulations
    with the Berger-POPC-07/H\"oltje-CHOL-13, CHARMM36 and MacRog force fields. The signs in the experimental values are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.  Most order parameters from Berger-POPC-07/H\"oltje-CHOL-13 model for g$_1$ are beoynd the y-axis scale.
    In order to elucidate the relative change as a function of cholesterol content,
    the simulation results are shifted such that the (smaller) $S_\mathrm{CH}$
    matches (within $\pm$0.01) the experimental value without cholesterol;
    the shift magnitudes for each of the force fields are listed ($S_\mathrm{CH}$+shift) in the y-label.
}
	\todo{Markus prepared two versions of this plot. The one on left has all the data points joint by lines;
	the one on right shows smooth trend lines (cubic polynomials) to ease qualitative comparison.
	Samuli likes the one on left, Markus the one on right. Do others have opinions?}
\end{figure}

In addition to the experimental data, the previously published simulation results from the Berger-POPC-07/H\"oltje-CHOL-13 model~\cite{ferreira13},
and our results from CHARMM36 and MacRog force fields are shown in Fig.~\ref{ordPchol}. 
To elucidate changes, the simulation results are shifted to match better with experiments without cholesterol.
The magnitude of shifting is written in the y-axis label.
As already pointed out previously, the Berger-POPC-07/H\"oltje-CHOL-13 model
seriously exaggerates the effect of cholesterol on the phospholipid glycerol backbone and choline segments~\cite{ferreira13}.
In comprasion, the choline responses of both CHARMM36 and MacRog are in better qualitative agreement with experiments,
and also the glycerol backbone responses appear less drastic.
%However, CHARMM36 seems to better reproduce the experimentally observed modest changes in the glycerol backbone segments 
%g$_2$ and g$_3$ with high concentrations of cholesterol. 
Thus we have calculated the glycerol backbone dihedral angle distributions
as a function of cholesterol in CHARMM36 (shown in Supplementary material) to resolve the cholesterol-induced structural changes. The only detectable change due to the
addition of cholesterol is the small decrease of gauche(-) and increase of gaughe(+) probability of the g$_3$-g$_2$-g$_1$-O(\textit{sn}-1) dihedral.
In conclusion, the results suggest that the significant effect of cholesterol on the lipid structures observed in simulations~\cite{zhu07,rog09,alwarawrah12} are overestimated by the model used in these studies, and the cholesterol only induces very small structural changes in the glycerol backbone.

It should be noted that the CHARMM36 force field parameters (dihedral potentials) for the glycerol backbone have been tuned to reproduce the correct order parameters at fully hydrated conditions~\cite{klauda10}. 
This procedure contains a risk of overfitting, which would manifest itself as wrong responses to changing conditions. 
According to our results, this tuning does not seem to lead to overfitting problems in the case of dehydration or lipid--cholesterol mixtures.

>>>>

\subsubsection{Cholesterol-containing phospholipid bilayer}
As cholesterol is abundant in biological membranes and
has been suggested to be an important player, for example, in domain formation~\cite{simons04,somerharju09},
phospholipid--cholesterol interactions have been extensively studied with theoretical~\cite{huang99,zhu07,rog09,alwarawrah12} and
experimental~\cite{brown78,marsh10,ferreira13,marsh13} methods.
It is widely agreed that cholesterol orders lipid acyl tails and thus decreases the area per molecule (condensing effect),
but its influence on the lipid headgroup and glycerol backbone remains debated~\cite{huang99,simons04,somerharju09}.
It has been suggested, for example, that the surrounding phospholipids shield cholesterol from exposure to water by 
reorienting their headgroups (``umbrella model'')~\cite{huang99} or that cholesterol acts as a spacer between the headgroups to increase 
their entropy and dynamics (``superlattice model'')~\cite{somerharju09}. 
%\todo{It has been suggested that we should use direct quotations from original papers here for clarity.}
%Both of these suggestions 
Molecular dynamics simulations have supported both
the umbrella~\cite{alwarawrah12} as well as the superlattice~\cite{zhu07} model,
in addition to suggesting specific interactions of cholesterol with the glycerol backbone~\cite{rog09}.
In these studies~\cite{zhu07,rog09,alwarawrah12}
the responses of the glycerol backbone and choline headgroup 
to increasing cholesterol content were not, however, compared to experiments.

Fig.~\ref{ordPchol} shows the responses of the choline headgroup ($\beta$ and $\alpha$)
order parameters of POPC (measured by $^1$H-$^{13}$C NMR~\cite{ferreira13}) and DPPC
($^{2}$H NMR~\cite{brown78}) to increasing cholesterol content.
Again, the two independent data sets agree very well:
Only very modest ($\Delta S_\mathrm{CH}<0.03$) changes occur in  the choline order parameters as cholesterol content
increases from 0 to 60\%.
The extreme sensitivity of the
high resolution $^{2}$H NMR experiments
is beautifully demonstrated by the measurable~\cite{brown78}
(but barely visible on the scale used in Fig.~\ref{ordPchol})
cholesterol-induced forking of the $\alpha$ order parameter.

We note that the modest ($\Delta S_\mathrm{CH}<0.02$ for g$_1$; $<0.04$ for g$_2$, g$_3$; see Fig.~\ref{ordPchol})
effects of cholesterol
on the glycerol backbone order parameters of POPC
measured by $^1$H-$^{13}$C NMR~\cite{ferreira13}
agree well with the results for phosphatidylethanolamine (PE) measured by $^{2}$H NMR~\cite{ghosh82}.
This further supports the ideas that the glycerol backbone structural behaviour is independent of the
headgroup composition~\cite{gally81} and that the headgroup structure is largely independent of the acyl chain region content unless
charges are present~\cite{scherer87}.
\begin{figure}[]
  \centering
  \includegraphics[width=7.3cm]{../DATAreportediINblog/cholesterolization_lp.pdf}
    \includegraphics[width=7.3cm]{../DATAreportediINblog/cholesterolization_smooth.pdf}
  \caption{\label{ordPchol}
    The effect of cholesterol content on the glycerol backbone and choline order parameters in experiments~\cite{brown78,ferreira13} and simulations
    with the Berger-POPC-07/H\"oltje-CHOL-13, CHARMM36 and MacRog force fields. The signs in the experimental values are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.  Most order parameters from Berger-POPC-07/H\"oltje-CHOL-13 model for g$_1$ are beoynd the y-axis scale.
    In order to elucidate the relative change as a function of cholesterol content,
    the simulation results are shifted such that the (smaller) $S_\mathrm{CH}$
    matches (within $\pm$0.01) the experimental value without cholesterol;
    the shift magnitudes for each of the force fields are listed ($S_\mathrm{CH}$+shift) in the y-label.
}
	\todo{Markus prepared two versions of this plot. The one on left has all the data points joint by lines;
	the one on right shows smooth trend lines (cubic polynomials) to ease qualitative comparison.
	Samuli likes the one on left, Markus the one on right. Do others have opinions?}
\end{figure}

In addition to the experimental data, Fig.~\ref{ordPchol} shows
our results for the CHARMM36 and MacRog force fields
and the previously published~\cite{ferreira13}
Berger-POPC-07/H\"oltje-CHOL-13 results.
Note that the simulation data are shifted vertically to ease comparison with experimental responses.
As previously pointed out~\cite{ferreira13}, the Berger-based model
seriously exaggerates the effect of cholesterol on the phospholipid glycerol backbone and choline headgroup.
In comparison, the choline responses of CHARMM36 and MacRog are in better qualitative agreement with experiments, and the glycerol backbone responses, 
although still somewhat larger than what is seen experimentally,
much less drastic.
Therefore,
to resolve the nature of cholesterol-induced structural changes,
we calculated from CHARMM36 the glycerol backbone dihedral angle distributions
at various cholesterol contents (Supplementary material): The only detectable change
is the small decrease of gauche(-) and increase of gaughe(+) probability of the g$_3$-g$_2$-g$_1$-O(\textit{sn}-1) dihedral.
In conclusion, our results suggest that the significant effects of cholesterol on lipid conformations observed in simulations~\cite{zhu07,rog09,alwarawrah12} are overestimated by the computational models used;
cholesterol only induces very small structural changes in the glycerol backbone.

Finally, it is important to note that the CHARMM36 force field parameters (glycerol backbone dihedral potentials)
have been tuned to reproduce the experimental order parameters at full hydration~\cite{klauda10}. 
This approach introduces a risk of overfitting, which would manifest itself as wrong responses to changing conditions. 
Interestingly, according to our results, tuning did not lead to overfitting problems as far as dehydration or cholesterol content are considered.
